---
date: last-modified
date-format: long
execute:
  freeze: auto  # re-render only when source changes
  cache: false
warning: false
message: false
eval: true
echo: false
---

# Repertoire analysis {.unnumbered}

```{r}
#| label: "setup"

library(magrittr)
library(tidyverse)
library(circlize)
library(ComplexUpset)
library(Polychrome)
library(ggridges)
library(yaml)

samples <- "../../samples.csv" |>
  vroom::vroom()
config <- "../../config/config.yaml"
config <- read_yaml(config)

groups <- config[["summary"]][["groups"]]

theme_set(theme_bw() +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(colour = "navyblue", size = 14)))

samples <- samples |>
  mutate(
    aux = glue::glue("../../output/qc/{sample_name}_trimmed.tsv"),
    aux = map(aux, readr::read_tsv),
    aux = map(aux, dplyr::select, -sample)) |>
  unnest(cols = c(aux)) |>
  mutate(
    diff = original - trimmed,
    original = scales::comma(original, 1),
    diff = scales::comma(diff, 1))

mc <- config[["summary"]][["min_count"]]
vj_quant <- config[["summary"]][["vj_quantile"]]
complete_str <- as.character(glue::glue("Complete >= {mc})", mc = mc))


# add repertoire summary statistics
if (!is.null(groups)) {

  samples <- map(groups, ~ filter(samples, sample_name %in% .x)) |>
    map2(names(groups), ~ mutate(.x, group = .y)) |>
    bind_rows()
}

samples <- samples |>
  mutate(
    aux = glue::glue("../../output/clonotypes/mixcr/{sample_name}/{sample_name}_airr.tsv"),
    clono = map(aux, readr::read_tsv,
      col_select = c(
        - ends_with("end"),
        - ends_with("start"),
        - contains("cigar"),
        - starts_with("np"),
        - starts_with("fwr"),
        - starts_with("cdr1"),
        - starts_with("cdr2"),
        - starts_with("junction"),
        - ends_with("alignment"))))

if ("group" %in% names(samples)) {

  samples <- samples |>
    group_by(sample_name, group)

} else {

  samples <- samples |>
    mutate(group = "all") |>
    group_by(sample_name, group)

}


samples <- samples |>
  unnest(cols = c(clono)) |>
  rename(count = duplicate_count)




```

## Count distribution


```{r}
#| label: "fig-count-dist"
#| include: true
#| fig-width: 10
#| fig-height: 6
#| out-width: "100%"
#| fig-cap: "# of clone distribution per sample"

out_plot <- samples |>
  ggplot(aes(y = count, x = sample_name)) +
    geom_violin(draw_quantiles = TRUE) +
    scale_y_log10(labels = scales::label_log()) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1))

if ("group" %in% names(samples)) {
  out_plot <- out_plot +
    facet_grid(cols = vars(group), scales = "free", space = "free_x")
}

out_plot +
  labs(x = "sample" , y = "# of clones")

```

## Shared clonotypes per group

::: {.callout-note collapse="true"}
The figures below have clonotypes repeated $\geq$ `r mc` times
:::

::: {.panel-tabset}

```{r}
#| label: "upset-setup"
#| include: false

get_upset_pdata <- function(clono, clono_var) {

  clono_enq <- syms(clono_var)

  pdata <- clono |>
    select(sample_name, count, !!! clono_enq) |>
    group_by(sample_name, !!! clono_enq) |>
    summarize(count = sum(count), .groups = "drop") |>
    pivot_wider(names_from = sample_name, values_from = count,
      values_fill = 0)

  mean_counts <- pdata |>
    select(where(is.numeric)) |>
    as.matrix() |>
    apply(1, FUN = function(x)mean(x[x > 0]))

  total_counts <- pdata |>
    select(where(is.numeric)) |>
    as.matrix() |>
    rowSums()

  count_breaks <- sort(c(10^{seq(0, 50, by = 1)}, 5))
  ww <- which(count_breaks  < max(total_counts))
  count_breaks <- c(count_breaks[count_breaks < max(total_counts)],
    Inf)

  lls <- map_chr(count_breaks, ~
    scales::comma(.x, scale = ifelse(.x < 1e3, 1, 1e-3),
      suffix = ifelse(.x < 1e3, "", "K")))
  lls <- str_c(lls[-length(lls)], lls[-1], sep = ", ")
  lls[-c(1, length(lls))] <- str_c("(", lls[-c(1, length(lls))])
  lls[-c(length(lls))] <- str_c(lls[-c(length(lls))], "]")
  lls[1] <- str_c("[", lls[1])
  last_val <- max(count_breaks[! is.infinite(count_breaks)])
  if (last_val >= 1e3) {
    last_ll <- scales::comma(last_val, 1, scale = 1e-3, suffix = "K")
  } else {
    last_ll <- scales::comma(last_val, 1)
  }
  lls <- c(lls[-length(lls)], str_c(">", last_ll, sep = " "))


  pdata <- pdata |>
    mutate(
      across(where(is.numeric), list(~ . > 0),
        .names = "{.col}")) |>
    mutate(
      mean_counts,
      total_counts,
      count_br = cut(total_counts, breaks = count_breaks,
        labels = lls,
        include.lowest = TRUE))

  pdata

}

upset_wrap <- function(clono, clono_var, df_label, n_inter = 17) {
  
  clono_enq <- syms(clono_var)

  pdata <- get_upset_pdata(clono, clono_var)
  
  vars <- pdata |>
    select(- one_of(clono_var)) |>
    select(- contains("count")) |>
    names()

  upset(pdata, vars,
    n_intersections = n_inter,
    sort_sets = FALSE,
    base_annotations = list(
      'Intersection size' = (
        intersection_size(
          text_mapping = aes(
            label = ifelse(
              !! get_size_mode("exclusive_intersection") < 1.1e3,
              paste0(round(!! get_size_mode("exclusive_intersection"), 1)),
              paste0(round(!! get_size_mode("exclusive_intersection") / 1e3, 1), "K"))),
          counts = TRUE) +
        scale_y_continuous(labels = scales::comma_format(.1, scale = 1e-3,
          suffix = "K")) +
        annotate("text", x = Inf, y = Inf, label = df_label,
            vjust = 1, hjust = 1)),
      "Expansion" = (
        ggplot(mapping = aes(fill = count_br)) +
          geom_bar(stat= "count", position = "fill") +
          scale_y_continuous(labels = scales::percent) + 
          scale_fill_viridis_d() +
          labs(y = "Expansion") +
          theme(legend.title = element_blank())),
      "Eff avg. counts" = (
        ggplot(mapping = aes(y = mean_counts)) +
          geom_violin(adjust = 2) +
          scale_y_log10(
            labels = scales::label_log(),
            breaks = scales::breaks_log(base = 10))
      )),
      set_sizes = FALSE, min_degree = 1)
          
}



render_upset <- function(group, group_lb, n_inter) {

  res = knitr::knit_expand(
    file = "template_upset.qmd",
    group = group,
    group_lb = group_lb,
    ninter = ninter)
  res

}

if ("group" %in% names(samples)) {
  groups <- unique(samples$group)
  group_lb <- str_c(groups, "samples", sep = " ")
} else {
  groups <- "all"
  group_lb <- "all samples"
}

ninter <- config[["report"]][["n_inter"]]
unparsed <- map2(groups, group_lb, render_upset, ninter)
parsed <- knitr::knit_child(text = unlist(unparsed),
  envir = rlang::env(
    samples = samples |>
      filter(count >= mc),
    ninter = ninter))

```

`r parsed`



:::

## Diversity analysis 

::: {.callout-note icon="false" collapse="true" title="Diversity definition"}
The diversity is generalized to the [Renyi entropy](https://en.wikipedia.org/wiki/R%C3%A9nyi_entropy) defined as:

$$
H(\alpha) = \frac{1}{1 - \alpha} \log \left( \sum_{i=1}^n p_i^\alpha  \right)
$$

where:

- $n$ is the total number of unique clonotypes
- $p_i$ is the clonotype frequency for clonotype $i$
:::

```{r}
#| label: "fig-diversity-curve"
#| fig-width: 12
#| fig-height: 6
#| out-width: "100%"
#| fig-cap: "Renyi entropy curves"

diversity_curve <- function(seqs, p, npoints = 100) {

  p <- p / sum(p)
  out <- tibble::tibble(alpha = seq(0, 2, length.out = npoints)) |>
    dplyr::mutate(
      div = map_dbl(alpha,
        ~ if (. == 0) {
          log(nrow(distinct(seqs)))
        } else if (. == 1) {
          - sum(p * log(p))
        } else {
          (1 / (1 - .)) * log(sum(p ^ .))
        }),
      div = div / log(nrow(distinct(seqs))))

  out

}



div_curves <- samples |>
    group_by(sample_name) |>
    reframe(
      diversity_curve(tibble(v_call, j_call, cdr3), count, npoints = 100)) |>
    mutate(
      clonality = 1 - div)

set.seed(935234)
npals <- length(unique(div_curves$sample_name))
pal <- createPalette(npals,
    c("#FF0000", "#00FF00", "#0000FF"), range = c(30, 80))
names(pal) <- unique(div_curves$sample_name)

div_curves |>
  ggplot(aes(x = alpha, y = div)) +
    geom_line(aes(colour = sample_name), linewidth = .75) +
    labs(x = expression(alpha), y = "normalized Renyi curve") +
    theme(legend.position = "top") +
    scale_color_manual(values = pal)
    
```

## VJ gene usage

```{r}
#| label: "fig-vj-usage"
#| include: true
#| fig-cap: "Gene frequency usage per group"
#| fig-width: 10
#| fig-height: 7
#| out-width: "100%"

vjgene_usage_barplot <- function(clono, clono_var) {


  clono_enq <- syms(clono_var)

  pdata <- clono |>
    group_by(sample_name, group, !!! clono_enq) |>
    summarize(
      count = sum(count), .groups = "drop")

  pdata_v <- pdata |>
    group_by(sample_name, group, v_call) |>
    summarize(count = sum(count), .groups = "drop") |>
    group_by(sample_name, group) |>
    mutate(freq = count / sum(count), type = "V") |>
    ungroup() |>
    rename(gene = v_call)

  pdata_j <- pdata |>
    group_by(sample_name, group, j_call) |>
    summarize(count = sum(count), .groups = "drop") |>
    group_by(sample_name, group) |>
    mutate(freq = count / sum(count), type = "J") |>
    ungroup() |>
    rename(gene = j_call)

  pdata <- bind_rows(pdata_v, pdata_j) |>
    group_by(gene) |>
    filter(!all(freq <= .01)) |>
    ungroup() |>
    mutate(
      gene = tidytext::reorder_within(gene, freq, type),
      type = factor(type, levels = c("V", "J")))

  pdata |>
    filter(! str_detect(gene, "\\,"))
}

vj_usage_plot <- samples |>
  vjgene_usage_barplot(c("v_call", "j_call", "cdr3")) |>
  filter(str_detect(gene, regex("^TR"))) |>
  ggplot(aes(y = gene, x =  freq)) +
    geom_boxplot() +
    tidytext::scale_y_reordered() +
    scale_x_continuous(labels = scales::percent) +
    labs(x = "Percentage of annotated clonotypes")    

if ("group" %in% names(samples)) {

  vj_usage_plot <- vj_usage_plot +
    facet_grid(rows = vars(type), cols = vars(group),
      scales = "free", space = "free")
} else {
  vj_usage_plot <- vj_usage_plot +
    facet_grid(rows = vars(type),
      scales = "free", space = "free")
}
    
vj_usage_plot   


```

### VJ genes paired usage


::: {.callout-note collapse="true"}
The figures use the clonotypes annotated by unique V and J genes, and ignoring the genes that portray less than `r scales::percent(vj_quant, 1)` of the total number of annotated clonotypes
:::

```{r}
#| label: "vj-setup"
#| include: false

get_vj_matrix <- function(repertoire, vcol = "V", jcol = "J") {

  stopifnot(
    vcol %in% names(repertoire),
    jcol %in% names(repertoire))

  # get match frequency by VJ genes
  out_mat <- repertoire |>
    filter(! str_detect(!!rlang::sym(vcol), ",")) |>
    filter(! str_detect(!!rlang::sym(jcol), ",")) |>
    group_by(!! rlang::sym(vcol), !! rlang::sym(jcol)) |>
    summarize(count = sum(count), .groups = "drop") |> 
    pivot_wider(names_from = all_of(jcol), values_from = "count",
      values_fill = 0) |>
      as.data.frame() |>
      column_to_rownames(vcol) |>
      as.matrix()
  out_mat

}





chord_diagram <- function(vj_mat, cols_v = NULL, cols_j = NULL,
  col_mat = NULL, title_str = NULL) {


  vgenes <- rownames(vj_mat)
  jgenes <- colnames(vj_mat)

  circos.clear()
  circlize::chordDiagram(vj_mat, grid.col = c(cols_v, cols_j),
    col = col_mat, annotationTrack = "grid", preAllocateTracks = 1)
  circos.trackPlotRegion(track.index = 1, panel.fun = function(x, y) {
    xlim <- circlize::get.cell.meta.data("xlim")
    ylim <- circlize::get.cell.meta.data("ylim")
    sector.name <- circlize::get.cell.meta.data("sector.index")
  circos.text(mean(xlim), ylim[1] + .1, sector.name, facing = "clockwise",
    niceFacing = TRUE, adj = c(0, 0.5))
  circos.axis(h = "top", labels.cex = 0.1, major.tick.length = 0.2,
    sector.index = sector.name, track.index = 2)
  }, bg.border = NA)

  if (! is.null(title_str)) {
    title(title_str)
  }

}

vj_mat_list <- samples |>
  group_split(sample_name) |>
  map(get_vj_matrix, "v_call", "j_call")

common_v <- map(vj_mat_list, rowSums) |>
  map(~ .[. > quantile(., vj_quant)]) |>
  map(names) |>
  reduce(intersect)

common_j <- map(vj_mat_list, colSums) |>
  map(~ .[. > quantile(., vj_quant)]) |>
  map(names) |>
  reduce(intersect)

vj_mat_list <- map(vj_mat_list, ~ .[common_v, common_j])

col_mat <- circlize::rand_color(length(common_v) * length(common_j),
  transparency = 0.3)
dim(col_mat) <- c(length(common_v), length(common_j))
rownames(col_mat) <- names(common_v)
colnames(col_mat) <- names(common_j)

set.seed(935234)
npals <- length(common_v)
pal_v <- createPalette(npals,
    c("#FF0000", "#00FF00", "#0000FF"), range = c(30, 80))
names(pal_v) <- common_v

pal_j <- Polychrome::alphabet.colors(length(common_j))
names(pal_j) <- common_j

names(vj_mat_list) <- group_split(samples, sample_name) |>
  map_chr( ~ .$sample_name[1])

render_chord <- function(sample) {

  res = knitr::knit_expand(
    file = "template_chord.qmd",
    sample_name = sample)
  res


}

vj_mat_list <- tibble(sample = names(vj_mat_list), vj_mat = vj_mat_list)

unparsed <- map(vj_mat_list$sample, render_chord)

parsed <- knitr::knit_child(text = unlist(unparsed),
  envir = rlang::env(
    vj_mat_list = vj_mat_list,
    col_mat = col_mat,
    pal_v = pal_v, pal_j = pal_j))


```

::: {.panel-tabset}

`r parsed`

:::


## CDR3 sequence length analysis


```{r}
#| label: "fig-cdr3-length"
#| fig-width: 9
#| fig-height: 8
#| out-width: "85%"
#| fig-cap: "Histograms of CDR3 AA sequences' length"

samples <- samples |>
  mutate(cdr3len = nchar(cdr3_aa)) 
  

samples |>
  ggplot(aes(x = cdr3len, y = sample_name)) +
    geom_density_ridges(stat = "binline", binwidth = 1, scale = 0.95,
      draw_baseline = FALSE) +
    facet_grid(rows = vars(group), scales = "free_y", space = "free_y") +
    labs(x = "CDR3 AA sequence length")

```

```{r}
#| echo: false
#| child: "appendix.qmd"

```
